---
title: "Cyclistic Member vs Casual Rider Analysis"
author: "John Judge"
date: '2022-06-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
## Identify how member and casual riders use cycles differently

## Give credit where credit is due:
This script follows the [Divvy Exercise R Script.](https://docs.google.com/document/d/1TTj5KNKf4BWvEORGm10oNbpwTRk1hamsWJGj6qRWpuI/edit)

It is NOT strictly copy/paste, but without the help provided by that script, I wouldn't have been able to do some of what's below.

## Load Libraries And Set Working Directory

```{r Load Libraries}
library("tidyverse")
library("lubridate")
library("anytime")
library("ggplot2")
```

```{r Set Working Directory}
setwd("C:/Users/jj/Documents/Google Data Analytics Cert/capstone project/Cyclistic Member vs Casual Rider Analysis")
```

## Set Up Data Frame

Read in Data From Separate Files

```{r }
data_202205 <- read.csv('202205-divvy-tripdata.csv')
data_202204 <- read.csv('202204-divvy-tripdata.csv')
data_202203 <- read.csv('202203-divvy-tripdata.csv')
data_202202 <- read.csv('202202-divvy-tripdata.csv')
data_202201 <- read.csv('202201-divvy-tripdata.csv')
data_202112 <- read.csv('202112-divvy-tripdata.csv')
data_202111 <- read.csv('202111-divvy-tripdata.csv')
data_202110 <- read.csv('202110-divvy-tripdata.csv')
data_202109 <- read.csv('202109-divvy-tripdata.csv')
data_202108 <- read.csv('202108-divvy-tripdata.csv')
data_202107 <- read.csv('202107-divvy-tripdata.csv')
data_202106 <- read.csv('202106-divvy-tripdata.csv')
```

Output Column Names For Each File To Ensure They are all the Same
```{r }
colnames(data_202205)
colnames(data_202204)
colnames(data_202203)
colnames(data_202202)
colnames(data_202201)
colnames(data_202112)
colnames(data_202111)
colnames(data_202110)
colnames(data_202109)
colnames(data_202108)
colnames(data_202107)
colnames(data_202106)
```

Combine Files Into Single Data Frame

```{r }
all_trips <- bind_rows(data_202205, data_202204, data_202203, data_202202, 
                       data_202201, data_202112, data_202111, data_202110,
                       data_202109, data_202108, data_202107, data_202106)
```

## Inspect Newly Created Data Frame

check columns, number of rows, structure and data types

```{r }
colnames(all_trips)
nrow(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

First attempt at view(all_trips) crashed RStudio!!!  No wonder, over 5M records :)

Create a slice (output not included, eval set to FALSE)
``` {r eval = FALSE}
all_trips_v0.1 <- slice(all_trips, 1:100)
view(all_trips_v0.1)
```

Inspect rideable_type and member_casual column values to ensure there are no slight variations in spelling/abbreviation throughout the dataset

```{r results = 'asis'}
unique(all_trips$rideable_type)
unique(all_trips$member_casual)
```

Inspect start and end station names for slight variations create a list to view since theres over 1000 unique values
Output not included, eval set to FALSE
```{r eval = FALSE}
start_station_names <- sort(unique(all_trips$start_station_name))
end_station_names <- sort(unique(all_trips$end_station_name))
view(start_station_names)
view(end_station_names)
```

There's probably a better way to do that.  This requires scrolling and looking for slight variants.

check ride_id column for duplicated IDs
```{r }
sum(duplicated(all_trips$ride_id))
```

## Add Fields to Enable Aggregating Data

add date related fields
```{r }
all_trips$date <- anydate(all_trips$started_at)
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$weekday <- format(as.Date(all_trips$date), "%A")
all_trips$month <- format(as.Date(all_trips$date), "%b")
all_trips$ridelength <- difftime(anytime(all_trips$ended_at), anytime(all_trips$started_at), units = "min")
```

create and view a slice of the data frame
Output not included, eval set to FALSE
```{r eval = FALSE}
all_trips_v1 <- slice(all_trips,1:100)
view(all_trips_v1)
```

convert ridelength column to numeric to enable calcs

```{r }
all_trips$ridelength <- as.numeric(all_trips$ridelength)
```

Inspect min, max, median and mean for ridelength column

```{r }
summary(all_trips$ridelength)
```

## Create A New Data Frame To Enable Analysis Of Valid Values

Exclude ride_length less than 1, since view() showed records with 0 mins and the summary() function showed min value is less than 0

NOTE - 0 min ride lengths do NOT have the same start and stop times!!!

ALSO - max value of 55944.15 (given a median of 11.13 and mean of 21.02) seems odd.  

Arbitrarily sub-setting data at this point for rides greater than 1 day and 1 week for closer look

### Create New Data Frames

data frame for all non-negative values
```{r }
all_trips_v2 <- all_trips[!(all_trips$ridelength < 1),]
```

create new data frame to view trips longer than 1 week
``` {r }
all_trips_v3 <- all_trips[(all_trips$ridelength > 10080),]
```

create new data frame to view trips longer than 1 day
``` {r }
all_trips_v4 <- all_trips[(all_trips$ridelength > 1440),]
```

Review rides greater than a day and week
Output not included, eval set to FALSE
``` {r eval = FALSE}
View(all_trips_v3)
View(all_trips_v4)
```

## Descriptive Analysis

Overall data set
```{r  }
summary(all_trips_v2$ridelength)
sd(all_trips_v2$ridelength)
hist(all_trips_v2$ridelength)
```

Max ride length of 55944.15 mins is troubling.  With only 4,338 rides greater than a day and 301 rides greater than a week, the mean is skewed slightly with these data points in the set.  Greater than a day seems like atypical riders, given there are 5M total records in the data set.

Pare down the data set to rides that aren't greater than a day and not less than 0
```{r }
all_trips_v2.1 <- all_trips[!(all_trips$ridelength > 1440) & !(all_trips$ridelength < 1),]
```

Review distribution of pared down data set
```{r }
summary(all_trips_v2.1$ridelength)
sd(all_trips_v2.1$ridelength)
hist(all_trips_v2.1$ridelength)
```

Still a significant spread in the data, 3rd quartile is only about 21 mins and the max is 1440 mins.  There is an incredibly long right tail in the distribution

## Calculate The Mild Outlier Thresholds
### Determine a suitable upper bound and possibly show a different group of ride characteristics based on ride length

Create a value so I can manually check the math the code is doing...
```{r }
ridelength_IQR <- IQR(all_trips_v2.1$ridelength)
```

Calculate lower and upper thresholds
```{r }
if(quantile(all_trips_v2.1$ridelength, probs = .25, na.rm = TRUE)-
  (1.5 * IQR(all_trips_v2.1$ridelength)) < 0)
  {
  lower_threshold <- 0
  }else {
    lower_threshold <- quantile(all_trips_v2.1$ridelength, probs = .25, na.rm = TRUE)-
      (1.5 * IQR(all_trips_v2.1$ridelength))
  }

upper_threshold <- quantile(all_trips_v2.1$ridelength, probs = .75, na.rm = TRUE) +
  (1.5 * IQR(all_trips_v2.1$ridelength))
```

Create data frames for both sides of the upper threshold to enable separate reviews
```{r }
all_trips_v2.2 <- all_trips_v2.1[(all_trips_v2.1$ridelength < 42.5),]
all_trips_v2.3 <- all_trips_v2.1[!(all_trips_v2.1$ridelength < 42.5),]
```

## Second Descriptive Analysis

Compare mean median min and max for casual vs member riders first for overall data set
```{r }
aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual, FUN = mean)
aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual, FUN = median)
aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual, FUN = min)
aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual, FUN = max)
```

Review rides by day of week, all rides

I WILL have order!
```{r }
all_trips_v2.1$weekday <- ordered(all_trips_v2.1$weekday, 
  levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual + all_trips_v2.1$weekday, FUN = mean)
```

Compare mean median max and min for data subset rides less than 42.5 mins
```{r }

aggregate(all_trips_v2.2$ridelength ~ all_trips_v2.2$member_casual, FUN = mean)
aggregate(all_trips_v2.2$ridelength ~ all_trips_v2.2$member_casual, FUN = median)
aggregate(all_trips_v2.2$ridelength ~ all_trips_v2.2$member_casual, FUN = min)
aggregate(all_trips_v2.2$ridelength ~ all_trips_v2.2$member_casual, FUN = max)
```

Review rides by day of week rides less than 42.5 mins
```{r }
all_trips_v2.2$weekday <- ordered(all_trips_v2.2$weekday, 
                                  levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2.2$ridelength ~ all_trips_v2.2$member_casual + all_trips_v2.2$weekday, FUN = mean)
```

Compare mean median max and min for data subset rides more than 42.5 mins
```{r }
aggregate(all_trips_v2.3$ridelength ~ all_trips_v2.3$member_casual, FUN = mean)
aggregate(all_trips_v2.3$ridelength ~ all_trips_v2.3$member_casual, FUN = median)
aggregate(all_trips_v2.3$ridelength ~ all_trips_v2.3$member_casual, FUN = min)
aggregate(all_trips_v2.3$ridelength ~ all_trips_v2.3$member_casual, FUN = max)
```

Review rides by day of week rides more than 42.5 mins
```{r }
all_trips_v2.3$weekday <- ordered(all_trips_v2.3$weekday, 
                                  levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2.3$ridelength ~ all_trips_v2.3$member_casual + all_trips_v2.3$weekday, FUN = mean)
```

## Create Plots And Visualize Data

Number of rides by day and by member type
```{r  echo = FALSE}
all_trips_v2.1 %>% 
  mutate(day_of_week = wday(anydate(started_at), label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_rides_v2.1 = n()) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides_v2.1, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Ride Count by Day of week
  and Rider Type", subtitle = "All Rides")
```

average ride length by day and by member type
``` {r echo = FALSE}
all_trips_v2.1 %>%
  mutate(day_of_week = wday(anydate(started_at), label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(average_duration_v2.1 = mean(ridelength)) %>% 
  arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = average_duration_v2.1, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Average Ride Duration by Day of week
  and Rider Type" ,subtitle = "All Rides")
```

Ride length by rideable type and member type
``` {r echo = FALSE}
all_trips_v2.1 %>%
  mutate(rideable_type, label = TRUE) %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(average_duration_v2.1 = mean(ridelength)) %>% 
  arrange(member_casual, rideable_type) %>% 
  ggplot(aes(x = rideable_type, y = average_duration_v2.1, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Average Ride Duration by Rideable Type and Rider Type", 
                                      subtitle = "All Rides")
```

Ride Count by Month and Rider Type
``` {r echo = FALSE}
all_trips_v2.1$month <- ordered(all_trips_v2.1$month, 
   levels=c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
            "Aug", "Sep", "Oct", "Nov", "Dec"))

all_trips_v2.1 %>% 
  mutate(month, label = TRUE) %>% 
  group_by(member_casual, month) %>% 
  summarize(number_of_rides_v2.1 = n()) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides_v2.1, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Ride Count by Month
  and Rider Type", subtitle = "All Rides")
```

## Executive Summary

**The question** - Identify how member and casual riders use cycles differently

**The Timeframe Analyzed** - June 2021 - May 2022

**Findings**

* Casual riders use bikes for longer periods of time
* Casual riders ride more than members on weekends
* Casual riders use “docked” bikes, members do not
* Casual Riders Use Bikes In Good Weather

**Recommendations**

* Identify repeat casual riders and/or demographics within this group
  + Create a focus group to determine why they don’t become members
* Aggregate data to enable a heat map of rides within the city
  + How are rider types distributed throughout the city
* Determine if there is an accessibility issue - identify an unmet need for casual riders
  + Since casual riders use docked bikes, whereas members do not
  + Casual riders ride longer - do they cover more distance or do they simply ride slower

## Appendix

### Rides can be as long as 55,944 mins

For reference, that’s longer than a month, which seems odd to be renting a bike
This led me down a path to find what drives riders to keep a bike for more than a day

* I found the upper threshold of ride times, 42.5 minutes (anything above this time is an outlier)

Number of rides by day and by member type (data set v2.2, rides less than 
42.5 minutes, the upper threshold for outliers in the overall data set)
``` {r echo = FALSE}
all_trips_v2.2 %>% 
  mutate(day_of_week = wday(anydate(started_at), label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_rides_v2.2 = n()) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides_v2.2, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Ride Count by Day of week
  and Rider Type", subtitle = "v2.2 - Rides less than 42.5 mins")
```

Average ride length by day and by member type
``` {r echo = FALSE}
all_trips_v2.2 %>%
  mutate(day_of_week = wday(anydate(started_at), label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(average_duration_v2.2 = mean(ridelength)) %>% 
  arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = average_duration_v2.2, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Average Ride Duration by Day of week
  and Rider Type" ,subtitle = "v2.2 - Rides less than 42.5 mins")
```

Number of rides by day and by member type (data set v2.3, rides more than 
42.5 minutes, the upper threshold for outliers in the overall data set)
``` {r echo = FALSE}
all_trips_v2.3 %>% 
  mutate(day_of_week = wday(anydate(started_at), label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_rides_v2.3 = n()) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides_v2.3, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Ride Count by Day of week
  and Rider Type",subtitle = "v2.3 - Rides more than 42.5 mins")
```

Ride Count by Rideable Type and Member Type
``` {r echo = FALSE}
all_trips_v2.1 %>% 
  mutate(rideable_type, label = TRUE) %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(number_of_rides_v2.1 = n()) %>% 
  arrange(member_casual, rideable_type)  %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides_v2.1, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Cyclistic Ride Count by Rideable Type
  and Rider Type", subtitle = "All Rides")
```

**Findings**

* Rides below 42.5 minutes are predominantly members, riding during the week
* Rides more than 42.5 minutes are predominantly casual, riding on weekends
  + They pick the bikes up on a weekend day…
* Rides more than 42.5 minutes average just above 90 minutes FOR BOTH GROUPS!
  + Casual average duration for this subset is 95 mins
  + Member average duration for this subset is 97 mins

**Takeaways and New Questions**

Takeaways

* Casual riders exhibit member characteristics and vice versa, depending on ride length
* Casual riders and members may use bikes for both commuting and recreation

New Questions

* Why haven’t certain casual riders become members?
  + Especially those where ride length is “short,” which suggests they are commuting
* Do members use bikes for commuting AND recreation???
  + Are members more likely to turn in bikes and take a new one during the same day?
  + Perhaps take a different bike for lunch break, etc, which would drive a longer ride length
* Do casual riders hang onto bikes longer because there is an extra step for them to get a different one?






## Create .csv files for visualizations in programs other than R
Output not included, eval set to FALSE

``` {r eval = FALSE}
ridecounts_day_type <- aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual + all_trips_v2.1$weekday, FUN = length)
write.csv(ridecounts_day_type, file = '~ride_count_day_type.csv')

ridelengths_day_type <- aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual + all_trips_v2.1$weekday, FUN = mean)
write.csv(ridelengths_day_type, file = '~avg_ride_length_day_type.csv')

ridelengths_month_type <- aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual + all_trips_v2.1$month, FUN = mean)
write.csv(ridelengths_month_type, file = '~ride_length_month_type.csv')

ridelengths_rideable_type <- aggregate(all_trips_v2.1$ridelength ~ all_trips_v2.1$member_casual + all_trips_v2.1$rideable_type, FUN = mean)
write.csv(ridelengths_rideable_type, file = '~ride_length_rideable_type.csv')
```

